Lower Bounds on the Minimax Risk for the \\ Source Localization Problem

### 2017-01-26 paper.tex
Make most final edits.

We give lower bounds on the minimax risk (mean squared-error in location) in
estimating the location of the source, which apply to all estimators, for
certain classes of diffusive media represented by low-pass
filters, when using a uniformly distributed sensor array. We show that
for sensors of a fixed size, the lower bound decays with
increasing numbers of sensors. We also perform a preliminary numerical analysis
to understand the effect of increasing the number of sensors to infinity by
shrinking their size, wherein the bound saturates for large sensor numbers.
In the second scenario, it is seen that there is greater
benefit to increasing the number of sensors as the signal-to-noise increases.
Our bounds are the first to give a scaling for the minimax risk in terms of the
number of sensors used.

@@

The source localization problem arises in many fields in different forms. Our
principal motivation comes from Electroencephalography (EEG), a non-invasive
brain measurement modality which uses electrodes placed on the scalp to sense
electric potentials produced by neuronal activity within the
brain~\cite{Nunez2006Electric}. The neurons (or groups of neurons) which
produce this activity are usually modeled as current dipoles, and it is often
of clinical or neuroscientific interest to estimate their
locations~\cite{Baillet2001Electromagnetic}. For a single activated dipole
within the brain (which is modeled as a point source), the electrodes
effectively sample a diffuse (blurred) representation of the dipole, as a
result of the spatial low-pass filtering effect of the different layers between
the brain surface and scalp surface (cerebrospinal fluid, skull and skin).

@@

Hence, we believe that this paper is the first to give lower bounds on source
localization error (measured in an $\ell_2$-distance sense) in estimating the
location of a one-dimensional point source, which is observed by sensors
through a diffusive medium (treated as a low-pass filter), and corrupted by
additive white Gaussian noise. We are also the first to analyze the
implications physical sensor models on such lower bounds. A simplistic
theoretical model might assume that the number of sensors can increase
unbounded, while maintaining a fixed SNR at each sensor. We show that this is
not physically possible in our setting, and derive numerical bounds which have
radically different asymptotic behaviour.

We describe the problem setting in detail in
Section~\ref{sec:source-localization} and the minimax techniques and our main
results (for a na\"ive and then for a more physical sensor model) in
Section~\ref{sec:minimax-lower-bounds}. We conclude in
Section~\ref{sec:discussion} with a discussion on the goodness of our bounds,
and on how our results can be extended to other simple settings.

@@

In the ``point sensor model'', sensors \emph{sample} the shifted
continuous-space impulse response, with some additive noise. We denote the
noiseless sampled version of $x(s; \theta)$ by the $m$-length vector $\v
x(\theta)$,
\begin{equation} \label{eq:sampled-signal}
	\v x(\theta) = \bigg[x(0; \theta), \ldots, x\Big(\frac{kS}{m}; \theta\Big), \ldots, x\Big(\frac{(m{-}1)S}{m}; \theta\Big)\bigg]^T
\end{equation}
where $m$ is the number of sensors and $k \in \{0, 1, \ldots, m-1\}$. Additive
noise $\v \epsilon$ corrupts $\v x(\theta)$ to produce the noisy sensor values
\begin{equation} \label{eq:sensor-obs}
	\v y = \v x(\theta) + \v \epsilon.
\end{equation}
The noise $\v \epsilon$ is zero-mean and Gaussian, $\v \epsilon \sim
\mathcal{N}(\v 0, \m \Sigma)$. We restrict our analysis to a ``sensor noise''
setting, wherein $\m\Sigma = \sigma^2 \m I$, i.e., noise is i.i.d.\ across
sensors.

In the ``integrator sensor'' model, sensors have a \emph{width} which decreases
as their number grows to infinity. Concretely, the $k$-th sensor which is
located at $s = kS/m$ \emph{occupies} the space $((k{-}\frac{1}{2})S/m,
(k{+}\frac{1}{2})S/m)$.  The signal sensed by this sensor is given by the
$k$-th element of $\v x(\theta)$,
\begin{equation}
	x_k(\theta) = \int_{(k-\frac{1}{2})S/m}^{(k+\frac{1}{2})S/m} x(s;\theta) ds.
\end{equation}
In this model, the noise vector $\v \epsilon$ is also generated similarly, by
integrating over small sections of a white noise process, which is assumed to
additively corrupt the continuous-space signal $x(s;\theta)$ \emph{prior} to
sampling (see Fig.~\ref{fig:signal-model}). This leads to i.i.d.\ noise values
at sensors (since sensors have no overlap), with variance inversely
proportional to $m$, i.e., $\sigma(m) \propto 1/\sqrt{m}$.

In the source localization problem, we receive $n$ ``trials''\footnote{The word
	``trials'' comes from neuroscience literature, where multiple trials of an
	experiment are run, with the same stimulus being given to the participant
	in each trial, evoking the same response in the brain (with some
trial-to-trial variability, which we ignore). For our purposes, we assume that
trials are effectively i.i.d.\ realizations of sensor noise.} (i.e., $n$
realizations) of $\v y$ according to equation~\eqref{eq:sensor-obs}, for the
\emph{same} point source $\theta$, using which we need to estimate the location
of the source.  The noise is also assumed to be i.i.d.\ across trials, i.e.,
the $n$ realizations of $\v y$ are independently drawn from $P(\theta)$, for
the same location parameter $\theta$.

@@

\begin{figure}[tp] %  figure placement: here, top, bottom, or page
	\centering
	\includegraphics[width=3.5in]{block-diagram}
	\caption{A diagrammatic representation of the signal model described in
	section~\ref{sec:signal-model}.}
	\label{fig:signal-model}
\end{figure}

@@

We performed a preliminary study to understand the effect of using the
``integrator sensor'' model defined in section~\ref{sec:sensor-model}.  The
bounding technique used in section~\ref{sec:lecam-lb} no longer easily admits a
closed-form solution which gives a bound for all values of $m$. We instead
compute the bound numerically, using exact expressions for error probability in
a binary hypothesis test. For Gaussian distributions of equal variance and a
uniform prior on the hypotheses, the minimum possible error probability (over
all tests) is a $Q$-function in the distance between their
means~\cite[Ch.~7]{Scharf1991Statistical},
\begin{equation}
	P_e = \inf_\psi \mathbb P(\psi(Y^n) \neq V) = Q\bigg(\frac{\norm{\v\mu_0 - \v\mu_1}}{2\sigma(m)}\bigg).
\end{equation}
where $Q(x) = \int_x^\infty \frac{1}{\sqrt{2\pi}} e^{-u^2/2} du$. Hence, from
Theorem~\ref{thm:est-to-testing},
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \Phi(\delta) P_e = \delta^2 Q\bigg(\frac{\norm{\v x(\theta_0) - \v x(\theta_1)}}{2\sigma(m)}\bigg).
\end{equation}
This bound is valid for any $\{\theta_0, \theta_1\}$ constituting a
$2\delta$-packing of $\Theta$. To numerically optimize over all possible
packings, we relax the process by fixing $\theta_0 = 0$ and optimizing over
$\theta_1 = 2\delta$.  We compute sensor values for a triangular impulse
response, by performing numerical integration, and find the best possible
$\delta$ by grid search.

\begin{figure}[t]
	\centering
	\includegraphics[width=2.8in]{lb-vs-m-annotated}
	\caption{A numerically generated plot of the lower bound on minimax risk of
		source localization, against the number of uniformly distributed
		sensors, $m$, used to estimate the location. The different curves are
		for different numbers of ``trials'', $n$ (number of i.i.d.\ sensor
		observations). The bound saturates for large $m$, indicating that for
		a \emph{fixed} number of trials, there is decreasing benefit to using
		more sensors.  However, with more trials, the bound saturates at
		increasingly larger numbers of sensors. Thus, there may be benefit to
	increasing the number of sensors when you have more trials.}
	\label{fig:numerical}
\end{figure}

Plotting this numerically computed bound for different numbers of trials (see
Fig.~\ref{fig:numerical}) reveals some interesting trends. First, the bound
saturates as $m$ grows large, matching the behaviour presented
in~\cite{Grover2016Fundamental}. Second, the lower bound on error is larger for
smaller numbers of sensors, as we expect. Third and most importantly, as the
number of trials increases, bound begins to saturate only for larger numbers of
sensors. In other words, there is increased benefit in using larger numbers of
sensors, when we also have more trials. This substantiates the intuition given
in~\cite{Grover2016Information}, which addresses this issue using a
Nyquist-rate perspective. A larger number of sensors will sample a larger
number of high-frequency components of the source signal (and hence allow us to
locate it better), but only if the noise floor is sufficiently low. Noise
variance, in turn, can be reduced by averaging over multiple trials.

@@

We also note that the minimax bounds from compressive sensing
literature~\cite{AriasCastro2013Fundamental} do not apply in our setting,
because those consider the problem of recovering the whole source signal,
$x(s;\theta)$. The loss function used there is hence the energy of the
difference signal $\int \abs{x(s) - \hat x(s)}^2 ds$. We, on the other hand,
are interested in bounding the minimax error in \emph{location}, i.e.,
$\abs{\theta - \hat\theta}^2$.

@@

Also note, the bounds derived in this paper apply only when sensors are
uniformly distributed.  Intuitively, this is the correct choice in the minimax
setting, since any non-uniform placement will have one or more points which is
far from all sensors. The ``maximization'' over $\theta \in \Theta$ would then
select a $\theta$ for which the value of the impulse response at the sensors is
smallest, so that SNR is minimized.

>>>>

We give lower bounds on the minimax risk (mean squared-error in location) in
estimating the location of the source, which apply to all estimators, for
certain classes of diffusive media, when using a uniformly distributed sensor
array. We show that for sensors of a fixed size, the lower bound decays with
increasing numbers of sensors. We also perform a preliminary numerical analysis
to understand the effect of increasing the number of sensors to infinity by
shrinking their size, wherein the bound saturates for large sensor numbers.  In
the second scenario, it is seen that there is greater benefit to increasing the
number of sensors as the signal-to-noise ratio increases.  Our bounds are the
first to give a scaling for the minimax risk in terms of the number of sensors
used.

@@

The source localization problem arises in many fields in different forms. Our
principal motivation comes from Electroencephalography (EEG), a non-invasive
brain measurement modality which uses electrodes placed on the scalp to sense
electric potentials produced by neuronal activity within the
brain~\cite{Nunez2006Electric}. The neurons (or groups of neurons) which
produce this activity are usually modeled as current dipoles, and it is often
of clinical or neuroscientific interest to estimate their
locations~\cite{Baillet2001Electromagnetic}. For a single activated dipole
within the brain (which is modeled as a point source), the electrodes sample a
diffuse (blurred) representation of the dipole, as a result of the spatial
low-pass filtering effect of the different layers of the skull, between the
brain and scalp.

@@

Hence, we believe that this paper is the first to give lower bounds on the
minimax error in estimating the location of a point source, which is observed
by sensors through a diffusive medium, and corrupted by additive white Gaussian
noise. We are also the first to analyze the implications physical sensor models
on such lower bounds. A simplistic theoretical model might assume that the
number of sensors can increase unbounded, while maintaining a fixed SNR at each
sensor. We show that this is not physically possible in our setting, and derive
numerical bounds which have radically different asymptotic behaviour.

We describe the problem setting in Section~\ref{sec:source-localization} and
the minimax techniques and our main results (for a na\"ive and then for a more
physical sensor model) in Section~\ref{sec:minimax-lower-bounds}. We conclude
in Section~\ref{sec:discussion} with a discussion on the goodness of our
bounds, and on how our results can be extended to other settings.

@@

In the ``point sensor'' model, sensors \emph{sample} the shifted
continuous-space impulse response, with some additive noise. We denote the
noiseless sampled version of $x(s; \theta)$ by the $m$-length vector $\v
x(\theta)$,
\begin{equation} \label{eq:sampled-signal}
	\v x(\theta) = \bigg[x(0; \theta), \ldots, x\Big(\frac{kS}{m}; \theta\Big), \ldots, x\Big(\frac{(m{-}1)S}{m}; \theta\Big)\bigg]^T
\end{equation}
where $m$ is the number of sensors and $k \in \{0, 1, \ldots, m-1\}$. Additive
noise $\v \epsilon$ corrupts $\v x(\theta)$ to produce the noisy sensor values
\begin{equation} \label{eq:sensor-obs}
	\v y = \v x(\theta) + \v \epsilon.
\end{equation}
The noise $\v \epsilon$ is zero-mean and Gaussian, $\v \epsilon \sim
\mathcal{N}(\v 0, \m \Sigma)$. We restrict our analysis to a ``sensor noise''
setting, wherein $\m\Sigma = \sigma^2 \m I$, i.e., noise is i.i.d.\ across
sensors.

In the ``integrator sensor'' model, sensors have a \emph{width} which decreases
as their number grows to infinity. Concretely, the \mbox{$k$-th} sensor which
is located at $s = kS/m$ \emph{occupies} the space $((k{-}\frac{1}{2})S/m,
(k{+}\frac{1}{2})S/m)$.  The signal sensed by this sensor is given by the
$k$-th element of $\v x(\theta)$,
\begin{equation}
	x_k(\theta) = \int_{(k-\frac{1}{2})S/m}^{(k+\frac{1}{2})S/m} x(s;\theta) ds.
\end{equation}
Here, a white noise process $\epsilon(s)$ additively corrupts the
continuous-space signal $x(s;\theta)$ prior to sampling (see
Fig.~\ref{fig:signal-model}). $\v \epsilon$ is generated by integrating
$\epsilon(s)$ over a sensor width. This leads to i.i.d.\ noise values at
sensors (since sensors have no overlap), with variance inversely proportional
to $m$, i.e., $\sigma(m) \propto 1/\sqrt{m}$ (refer
Appendix~\ref{app:integrator-snr} in the full version of this
paper~\cite{FullVersion} for an explanation).

@@

In the source localization problem, we receive $n$ ``trials''\footnote{The word
	``trials'' comes from neuroscience literature, where multiple trials of an
	experiment are run, with the same stimulus being given to the participant
	in each trial, evoking the same response in the brain (with some
trial-to-trial variability, which we ignore). For our purposes, we assume that
trials are i.i.d.\ realizations of sensor noise.} of $\v y$ according to
equation~\eqref{eq:sensor-obs}, for the \emph{same} point source $\theta$,
using which we need to estimate the location of the source. Note that noise is
assumed to be i.i.d.\ across trials, i.e., the $n$ realizations of $\v y$ are
independently drawn from $P(\theta)$, for the same location parameter $\theta$.

\begin{figure}[tp] %  figure placement: here, top, bottom, or page
	\centering
	\includegraphics[width=3.2in]{block-diagram}
	\caption{A block diagram of the two sensor models described in
	section~\ref{sec:signal-model}.}
	\label{fig:signal-model}
\end{figure}

@@

This observation motivates the ``integrator sensor'' model defined in
section~\ref{sec:sensor-model}. With this model, the bounding technique used in
section~\ref{sec:lecam-lb} no longer easily admits a closed-form solution which
gives a bound for all values of $m$. We instead compute the bound numerically,
using exact expressions for error probability in a binary hypothesis test. For
Gaussian distributions of equal variance and a uniform prior on the hypotheses,
the minimum possible error probability (over all tests) is a $Q$-function in
the distance between their means~\cite[Ch.~7]{Scharf1991Statistical},
\begin{equation}
	P_e = \inf_\psi \mathbb P(\psi(Y^n) \neq V) = Q\bigg(\frac{\norm{\v\mu_0 - \v\mu_1}}{2\sigma(m)}\bigg).
\end{equation}
where $Q(x) = \int_x^\infty \frac{1}{\sqrt{2\pi}} e^{-u^2/2} du$. Hence, from
Theorem~\ref{thm:est-to-testing},
\begin{equation}
	\mathfrak{M}_n(\mathcal{P}, \Phiorho) \geq \Phi(\delta) P_e = \delta^2 Q\bigg(\frac{\norm{\v x(\theta_0) - \v x(\theta_1)}}{2\sigma(m)}\bigg).
\end{equation}
This bound is valid for any $\{\theta_0, \theta_1\}$ constituting a
$2\delta$-packing of $\Theta$. To numerically optimize over all possible
packings, we relax the process by fixing $\theta_0 = 0$ and optimizing over
$\theta_1 = 2\delta$.  We compute sensor values for a triangular impulse
response, by performing numerical integration, and find the best possible
$\delta$ by grid search.

\begin{figure}[t]
	\centering
	\includegraphics[width=2.8in]{lb-vs-m-annotated}
	\caption{A numerically generated plot of the lower bound on minimax risk of
		source localization, against the number of uniformly distributed
		sensors, $m$, used to estimate the location (code available
		online~\cite{Code}). The different curves are for different numbers of
		``trials'', $n$ (number of i.i.d.\ sensor observations). The bound
		saturates for large $m$, indicating that for a \emph{fixed} number of
		trials, there is decreasing benefit to using more sensors.  However,
	with more trials, the bound saturates at increasingly larger numbers of
sensors. Thus, there may be benefit to increasing the number of sensors when
you have more trials.}
	\label{fig:numerical}
\end{figure}

Plotting this numerically-computed bound for different numbers of trials
reveals some interesting trends (see Fig.~\ref{fig:numerical}). First, the
bound saturates as $m$ grows large, matching the behaviour of the
information-theoretic bound. Second, the lower bound on error is larger for
smaller numbers of sensors, as we expect. Third and most importantly, as the
number of trials increases, the bound begins to saturate only at a larger
number of sensors. In other words, there is \emph{increased} benefit in using
larger numbers of sensors, when we \emph{also} have more trials. This
substantiates the intuition given in~\cite{Grover2016Information}, which makes
this point using a Nyquist-rate argument: a larger number of sensors will
sample a larger number of high-frequency components of the source signal (and
hence allow us to locate it better), but only if the noise floor is
sufficiently low. Noise variance, in turn, can be reduced by averaging over
multiple trials.

@@

We also note that the minimax bounds from compressive sensing
literature~\cite{AriasCastro2013Fundamental} \emph{do not} apply in our
setting, because those consider the problem of recovering the \emph{whole}
source signal, $x(s;\theta)$. The loss function used there is hence the energy
of the difference signal, $\int \abs{x(s) - \hat x(s)}^2 ds$. We, on the other
hand, are interested in bounding the minimax error in \emph{location}, i.e.,
$\abs{\theta - \hat\theta}^2$.

@@

Also note, the bounds derived in this paper apply only when sensors are
uniformly distributed.  Intuitively, this is the \emph{correct} choice in the
minimax setting, since any non-uniform placement will have one or more points
which is far from all sensors. The ``maximization'' over $\theta \in \Theta$
would then select a $\theta$ for which the value of the impulse response at the
sensors is smallest, so that SNR is minimized.
